Do professional chess players draw too often?

### 2022-03-08 paper/paper.tex
final? changes

\section*{Disclaimer}
I spent most of my time trying to get the models working. It was a complete nightmare. I thought it was close to working for many hours. Last 1\% took 99\% of the time.
Getting the factor analysis model to work (replacing Smerdon's PCA) took me about 10 hours alone.
Maybe I should have settled for PCA for this draft.
Perhaps I should not have tried to relearn Pyro, a probabilistic programming language written in Python.
That being said, if I had to build this model in R I likely would have ended up using RStan, which is a binding to another PPL, so the model wouldn't have really been in R either.
The data cleaning and manipulation would also have taken me 10x longer since I don't really know R.
Hindsight is 20/20. I think it's on its way, though.

The code (and this writeup) is in a rough state as a result -- by the final draft it will be pretty and commented and clear (despite being in Python).
Similarly, since the model was giving me issues I didn't have compute time to train the model on the whole dataset (which is 4.6M games after cleaning).
Again this will not be the case by the final draft. Everything works though; I have a complete pipeline
that just needs some clean up, some hyperparameters to be adjusted/played with, and to be run to completion.

Also, there are a couple of things I would like to add (or at least investigate) by the final draft:
\begin{itemize}
    \item Dataset of online games (ostensibly less incentive to draw here, more ``naturalistic'' or aggressive gameplay)
    \item Using the factor model to compute an index, comparing to Smerdon's results
\end{itemize}

@@

\section{Conclusion}
These results relate back to the article/real-world situation in a couple of ways.
First, it shows that variance-lowering behavior at the cost of expected value does occur in chess.
That is, playing suboptimally for draws at higher levels of play occurs often.
It is difficult to identify, however, if this is as a result of random noise or variance in playing strength
on behalf of players, or if each choice of $\alpha_W, \alpha_B$ for each game is intentional.
It is also clear that this behavior, on average, decreases as the level of play increases, as predicted.

@@

The article mentions measures designed to increase errors and decrease the likelihood of draws.
These would likely be successful at the highest echelons of play, where games are sufficiently
scrutinized and play is so precise that suboptimal play essentially does not exist.
However, at slightly lower levels of play (still professional, but just among weaker players), the problem is not absolute precision.
Rather, (assuming the results support this), the problem is this variance-reducing behavior.
So from a design perspective, the solution would be to penalize this behavior.
Making draws worth $\frac{1}{3}$ of a point could discourage taking draws in tournaments, for instance.
Removing agreed draws (where players simply agree to a draw rather than playing it out) as a whole could also make this behavior more difficult.

% TODO BIB
Caissabase, Elo, Smerdon, The Fuzz, original article
\end{document}

>>>>

\section*{Description of changes}
I changed all the models to be in principally game-theoretic terms.
The models were rewritten, much of the code has changed, but the basic ideas
and underlying assumptions have remained.
The Elo rating system is still used, but I also modified it to allow it to represent
suboptimal play and inform the predictive logistic regressions that are described later.

The only piece of feedback I was unable to incorporate was running simulations to generate
data (though I could have done this after training the regressions, it would not have been useful).
This is principally because of the way the Elo system works, and its description of an expected
value rather than as probabilities of winning, drawing, or losing.
So without making baseless assumptions (that would have resulted in useless simulations), generating
simulated data would not have been useful.
There was also plenty of real data to work with as well.

@@

\section{Conclusion}
These results relate back to \href{https://fivethirtyeight.com/features/some-humble-suggestions-to-save-chess-from-itself/}{Roeder's article}/the real-world situation in a couple of ways.

First, they shows that variance-lowering behavior at the cost of expected value does occur in chess.
That is, playing suboptimally for draws at higher levels of play occurs often.
It is difficult to identify, however, if this is as a result of random noise or variance in playing strength
on behalf of players, or if each choice of $\alpha_W, \alpha_B$ for each game is intentional.
It is also clear that this behavior, on average, decreases as the level of play increases, as predicted.

@@

Roeder mentions a few measures designed to decrease drawing behavior, but these again are for
the highest echelons (super GM levels) of play.
These measures increase errors and decrease the likelihood of draws, whether by shortening time
controls, forcing tiebreakers to be played before matches (to force playing for wins), and so on.
These would likely be successful at the highest levels of play, where games are sufficiently
scrutinized and play is so precise that suboptimal play essentially cannot exist.

However, at slightly lower levels of play (still professional, but just among weaker players), the problem is not absolute precision.
Rather, (assuming the results support this), the problem is this variance-reducing behavior.
So from a design perspective, the solution would be to penalize this behavior.
Making draws worth $\frac{1}{3}$ of a point could discourage taking draws in tournaments, for instance.
Removing agreed draws (where players simply agree to a draw rather than playing it out) as a whole could also make this behavior more difficult.

\begin{thebibliography}{9}

\bibitem{caissabase}
\href{http://caissabase.co.uk/}{Caissabase}. Unknown author.

\bibitem{elo}
Elo, Arpad (1986). \textit{The Rating of Chessplayers, Past and Present}.

\bibitem{polgar}
Polgar, Susan (2011). \href{https://chessdailynews.com/the-700-lbs-gorilla-issue-to-draw-or-not-to-draw/}{The 700 lbs gorilla issue: To draw or not to draw.} \textit{Chess Daily News}.

\bibitem{538}
Roeder, Oliver (2021). \href{https://fivethirtyeight.com/features/some-humble-suggestions-to-save-chess-from-itself/}{Some Humble Suggestions to Save Chess From Itself}. \textit{538}.

\bibitem{thefuzz}
SeatGeek (2022). \href{https://github.com/seatgeek/thefuzz}{TheFuzz}, Python package.

\bibitem{fci}
Smerdon, David (2021). \href{https://www.davidsmerdon.com/?p=2168}{\textit{The Fighting Chess Index}}.

\end{thebibliography}
\end{document}
